Interactive Visualization of Large Bipartite Networks Assisted by Multilevel Strategies\thanks{Supported by FAPESP, project 2017/05838-3}

### 2019-05-28 paper.tex
feita primeira figura
terminada primeira versao do artigo

\keywords{Network visualization \and Bipartite networks \and Multilevel strategies \and Big data \and Complex networks \and Data visualization.}
\end{abstract}
%
\section{Introduction}
The visualization of large-scale networks poses challenges both in terms of computational costs
and of effective presentation of the information to the user~\cite{tang,staudt}.
These issues may be aggravated in the case of bipartite networks,
due to their sparsity and topological complexities~\cite{alan1}.
Bipartite networks are comprised of two partitions of nodes,
called ``layers'',
and links are not incident between nodes in the same partition.
Such network type arises very often and naturally from the representation
of relations among two kinds objects,
e.g. documents and terms or authors~\cite{doc,sci,movie}, or patient and gene~\cite{gene}.
Furthermore, real-world networks are often bipartite, and most unipartite networks
are projections of bipartite networks or exhibit bipartite properties~\cite{guillaume0,guillaume}.
In order to assist the visualization and navigation of large networks, one possibility is the use of
multilevel strategies, which consist on the employment of incremental coarsening of the original
network to obtain a sequence of simplified representations, i.e. with fewer nodes and links.
Multilevel strategies are most traditionally used for executing complex algorithms
on large-scale networks by the application of the algorithm on a smaller
version of the network~\cite{alan2,ml2}.
The employment of multilevel strategies for the visualization of simple (i.e. ``unipartite'')
networks has been reported, but their exploitation for visualizing bipartite networks
was not found by the authors, as described in the next subsection.
Accordingly, we present a system for the visualization of bipartite networks using
multilevel strategies developed for bipartite networks.
The system consists on presenting a simplified version of the network to the user, which then
requests for supervertices (or collections of them) to be uncoarsened and visualized in more detail.
Thus, the resulting interface allows for the visualization of large structures;
enables an interactive navigation of the structure e.g. by uncoarsening supernodes, zooming and requesting information about the nodes;
and may assist not only in the analysis of bipartite networks, but also on the analysis and development of multilevel algorithms in bipartite networks.
Most importantly, the overall method, and the implemented interface, is a proof-of-concept, it demonstrates the feasibility of the use of multilevel strategies for the efficient visualization of large bipartite networks.

@@

A multilevel strategy consists in obtaining a hierarchy of coarsened networks $G_l$, $l$ integer and $l \in [0,L-1]$ where $G_0$ is the original network
and where $|V_i| \geq |V_{i+1}|$.
The coarsening procedure requires two algorithms, the \emph{matching}, that defines the nodes to be collapsed, and the \emph{contraction}, which builds the reduced representation
from the matched nodes.
There are several coarsening algorithms reported in the scientific literature~\cite{}, a few of them
developed for bipartite networks~\cite{}.
A detailed exposition of these algorithms is found on the bibliography~\cite{} and is beyond the scope of this article.
Most importantly, we are interested in coarsening suited for bipartite networks,
as it has yielded simplified networks which present essential topological features
of the original networks,
and such result is perceived in the visualization as hinted in~\cite{alan2} and further shown in
Section~\ref{sof}.

Visual analytics is the scientific field dedicated to analytical reasoning assisted by interactive visual interfaces~\cite{}.
Therefore, the area is specially concerned with coupling interactive visual representations with sense and decision making.
Of special relevance for the present work, the techniques are most often employed to amplify human capabilities in specific ways,
which includes reducing the search space, enhancing the recognition of patterns and the inference of relationships,
and providing a manipulable medium for the exploration of the information of interest~\cite{}.
For this present work to be better characterized as a visual analytics contribution, the methods described in this section and software described in Section~\ref{sof} should further enable the discovery of information, such as metadata of the networks (e.g. authors, words or documents).
Accordingly, our methods and software may be understood as data visualization with explicit visual analytics elements, such as interactivity, on-demand supply of information, and selective navigation.

@@

\begin{figure}[!h]\centering
 \includegraphics[width=0.6\textwidth]{fnav}
  \caption{The navigation pathway of the bipartite network obtained through a multilevel strategy
  and auxiliary tools.
  Compliant with the Shneiderman's mantra, first and overview of the bipartite network is provided.
  Then the user may select an specific portion of the network to visualize in further detail,
  apply filters and visual mappings of topological features, and request details.
  These operations may be carried on in any order and any number of times the analyst understand convenient.
  }\label{fnav}
\end{figure}

\noindent 
\section{Software implementation}\label{sof}
% optimization of the computational capacity for large networks by
% means of Web GL, triangles for nodes and details on demand
We implemented the framework described in Section~\ref{des} using scientific,
database and web resources,
and made the final result available to the user within a web page
for use through simple mouse-driven actions
and requiring the instalation of no software but a web browser such as Firefox or Google Chrome.
The final software was named MlBiNetVis (from Multilevel Bipartite Network Visualization),
is exemplified in Figures~\ref{fpage0} and~\ref{fpage}.
% TTM
Next subsections describe its functionalities and the technologies used.

\begin{figure}[!h]\centering
 \includegraphics[width=\textwidth]{fpage0}
  \caption{The navigation pathway of the bipartite network obtained through a multilevel strategy
  and auxiliary tools.
  }\label{fpage0}
\end{figure}

\begin{figure}[!h]\centering
 \includegraphics[width=\textwidth]{fpage}
  \caption{The navigation pathway of the bipartite network obtained through a multilevel strategy
  and auxiliary tools.
  }\label{fpage}
\end{figure}

\subsection{The use of MlBiNetVis}
% network upload, format.
% layouts available
% algorithms available for coarsening, and their parametrization
% initialization of the visualization: click on render
% tools available and their operation
In using MlBiNetVis,
the analyst first selects a network of interest to be explored.
One may also upload a network and then select it on the same drop-down menu.
The user may then wish to select one of the available network layout
algorithms and tune the coarsening to be performed, but default settings are reasonable
for the newcommer.
Upon request given by hiting the ``render network'' button,
the network is mapped to the screen.
Subsequent usage relies on manipulation of the visual mapping presented on the
canvas by means of actions
performed through managing the tools and related mouse operations.
These steps are illustrated in Figure~\ref{initial}.

% expose about text area, show links checkbox, layout button, and other elements. TTM

There are multiple levels of representation entailed by the multilevel strategy.
This makes pertinent that one and only of the available levels is explicitly selected
so that the transformations are performed on that level.
Thus, there is always a selected level in which the tools perform actions, such
as link coloring or opening supernodes to expose their predecessor nodes.
As can be infered by the reader,
the number of nodes and links in each level which are visible are not fixed thoughout the
navigation.
The selection of the level to which the transformations should apply, and the number
of visible and total nodes and links in each level, are exposed in the table to the right
of the canvas, as in Figure~\ref{secondPhase}.
In the same table, the user may change the color and shape of the nodes in each layer of
a level, the color of the links of a level, and toggle show/hide of the links is a level,
by left/right clicks on the corresponding colored cell.
These operations are illustrated in Figure~\ref{using}.

@@

The third group of buttons is more specific to the needs of the MlBiNetViz, and
are selected to make mouse operations entail special processes and transformations of the visualization.
The first button is selected for the interface to return core information from a clicked node:
id, level, number of predecessors of previous level, id of successor (if node is not on the last level),
number of neighbors (degree), sum of weight of links (strength).
Other information the analysts report necessity of are easily implemented, but currently only these informations
are returned to avoid unecessarily flooding the text area.
The second tool button is seleted to move nodes: specific nodes are individually moved if clicked and dragged;
a rectangular region is made available for moving all nodes therein if an empty region of the canvas
is clicked and the mouse cursor dragged before releasing the button.
The third tool button is selected for opening supernodes and exposing their predecessor nodes and links among them:
a region is defined by the user by clicking on an empty space on the canvas and gragging the mouse before releasing,
the supernodes therein are joined into a rectangular region that contains all of them,
and their predecessor nodes (which are on a finer level of detail) are drawn inside such rectangular region.
Open supernodes may be joined using the same tool, but it was found convenient to make available a
fourth button to assist in this task: if this tool is selected, one may join two open supernodes by
clicking on them in sequence.
The last tool button in this group is selected to perform three operations on open supernodes:
if an open supernode is clicked on and dragged, the rectangular shape is resized accordingly;
if there is no drag, i.e. the super node is clicked on and released, the links from it to the other
supernodes are attached to the next counterclockwise corner of its rectagular region.
When the open supernode is resized, the layout of its child nodes is optimized for the drawn region.
This is most useful when used in combination with the drag tool:
on the finer level of detail, the nodes are repositioned into a more compact disposition.
Then upon a (potentially minimal) supernode
risize, the child nodes are made to fit the whole rectangular area.
The last set of tools acts on the whole canvas:
changing the color of the background, zoom, left-right/up-down pan,
rotation of the entire network,
and toggling between current zoom and pan status and the initial setting.
The toolbar is contextualized in Figure~\ref{using} and their use is illustrated in Figure~\ref{tools}.

@@

This is a very brief account of the styles and software resources used to implement
MlBiNetVis. The interested reader is welcome to browse the software code and documentation
and contact the authors~\cite{mbvncode}.

@@

>>>>

\keywords{Network visualization \and Bipartite networks \and Multilevel strategies \and Big data \and Complex networks \and Data visualization.}
\end{abstract}
%
\section{Introduction}
The visualization of large-scale networks poses challenges both in terms of computational costs
and of effective presentation of the information to the user~\cite{tang,staudt}.
These issues may be aggravated in the case of bipartite networks,
due to their sparsity and topological complexities.
Bipartite networks are comprised of two partitions of nodes,
called ``layers'',
and links are not incident between nodes in the same partition.
Such network type arises very often and naturally from the representation
of relations among two kinds objects,
e.g. documents and terms or authors~\cite{doc,sci,movie}, or patient and gene~\cite{gene}.
Furthermore, real-world networks are often bipartite, and most unipartite networks
are projections of bipartite networks or exhibit bipartite properties~\cite{guillaume0,guillaume}.
In order to assist the visualization and navigation of large networks, one possibility is the use of
multilevel strategies, which consist on the employment of incremental coarsening of the original
network to obtain a sequence of simplified representations, i.e. with fewer nodes and links.
Multilevel strategies are most traditionally used for executing complex algorithms
on large-scale networks by the application of the algorithm on a smaller
version of the network~\cite{alan2,ml2}.
The employment of multilevel strategies for the visualization of simple (i.e. ``unipartite'')
networks has been reported, but their exploitation for visualizing bipartite networks
was not found by the authors, as described in the next subsection.
Accordingly, we present a system for the visualization of bipartite networks using
multilevel strategies developed for bipartite networks.
The system consists on presenting a simplified version of the network to the user, which then
requests for supervertices (or collections of them) to be uncoarsened and visualized in more detail.
Thus, the resulting interface allows for the visualization of large structures;
enables an interactive navigation of the structure e.g. by uncoarsening supernodes, zooming and requesting information about the nodes;
and may assist not only in the analysis of bipartite networks, but also on the analysis and development of multilevel algorithms in bipartite networks.
Most importantly, the overall method, and the implemented interface, is a proof-of-concept, it demonstrates the feasibility of the use of multilevel strategies for the efficient visualization of large bipartite networks.

@@

A multilevel strategy consists in obtaining a hierarchy of coarsened networks $G_l$, $l$ integer and $l \in [0,L-1]$ where $G_0$ is the original network
and where $|V_i| \geq |V_{i+1}|$.
The coarsening procedure requires two algorithms, the \emph{matching}, that defines the nodes to be collapsed, and the \emph{contraction}, which builds the reduced representation
from the matched nodes.
There are several coarsening algorithms reported in the scientific literature~\cite{co1,co2}, a few of them
developed for bipartite networks~\cite{alan2}.
A detailed exposition of these algorithms is found on the bibliography and is beyond the scope of this article.
Most importantly, we are interested in coarsening suited for bipartite networks,
as it has yielded simplified networks which present essential topological features
of the original networks,
and such result is perceived in the visualization as hinted in~\cite{alan2} and further shown in
Section~\ref{sof}.

Visual analytics is the scientific field dedicated to analytical reasoning assisted by interactive visual interfaces~\cite{visAn}.
Therefore, the area is specially concerned with coupling interactive visual representations with sense and decision making.
Of special relevance for the present work, the techniques are most often employed to amplify human capabilities in specific ways,
which includes reducing the search space, enhancing the recognition of patterns and the inference of relationships,
and providing a manipulable medium for the exploration of the information of interest.
For this present work to be better characterized as a visual analytics contribution, the methods described in this section and software described in Section~\ref{sof} should further enable the discovery of information, such as metadata of the networks (e.g. authors, words or documents).
Accordingly, our methods and software may be understood as data visualization with explicit visual analytics elements, such as interactivity, on-demand supply of information, and selective navigation.

@@

\begin{figure}[!h]\centering
 \includegraphics[width=0.8\textwidth]{fnav___}
  \caption{The navigation pathway of bipartite networks obtained through a multilevel strategy
  and auxiliary tools.
  Compliant with the Shneiderman's mantra, first and overview of the bipartite network is provided.
  Then the user may select an specific portion of the network to visualize in further detail,
  apply filters and visual mappings of topological features, and request details.
  These operations may be carried on in any order and any number of times the analyst understand convenient.
  }\label{fnav}
\end{figure}

\noindent 
\section{Software implementation}\label{sof}
% optimization of the computational capacity for large networks by
% means of Web GL, triangles for nodes and details on demand
We implemented the framework described in Section~\ref{des} using scientific,
database and web resources,
and made the final result available to the user within a web page
for use through simple mouse-driven actions
and requiring the instalation of no software but a web browser such as Firefox or Google Chrome.
The final software was named MlBiNetVis (from Multilevel Bipartite Network Visualization)
and is exemplified in Figures~\ref{fpage0},~\ref{initial} and~\ref{secondPhase}.
Next subsections describe its functionalities and the technologies used.

\begin{figure}[!h]\centering
 \includegraphics[width=\textwidth]{overall____}
  \caption{The MlBiNetVis interface. (1) is the dropdown menu for choosing the network to analyse or uploading new ones.
  The gray box in (2) contains widgets to select and parametrize the coarsening algorithm.
  (3) is the render button, which maps the network to the screen.
  In (4) are auxiliary widgets for selecting the  layout, show/hide of links, and for the interface to output information requested by the user or useful to assist navigation.
  In (5) is an interactive table which holds information about the levels and layers of the bipartite network and may be used to set the visualization on the canvas.
  In (6) is the toolbar that enables the navigation of the network and fine-tuning the visualization.
  In (7) is the canvas in which the network is drawn through node-link diagrams.
  Section~\ref{use} describes the use of these elements in further detail.
  }\label{fpage0}
\end{figure}

\subsection{The use of MlBiNetVis}\label{use}
% network upload, format.
% layouts available
% algorithms available for coarsening, and their parametrization
% initialization of the visualization: click on render
% tools available and their operation
In using MlBiNetVis,
the analyst first selects a network of interest to be explored.
One may also upload a network and then select it on the same drop-down menu.
The user may then wish to select one of the available network layout
algorithms and tune the coarsening to be performed, but default settings are reasonable
for the newcommer.
Upon request given by hiting the ``render network'' button,
the network is mapped to the screen.
Subsequent usage relies on manipulation of the visual mapping presented on the
canvas by means of actions
performed through managing the tools and related mouse operations.
These steps are illustrated in Figure~\ref{initial}.
\begin{figure}[!h]\centering
 \includegraphics[width=\textwidth]{initial_}
  \caption{The elements in the initial step of the MlBiNetVis usage.
  The button at the top is for network upload and selection.
  In the gray box, the user chooses and sets the coarsening algorithm as detailed in Appendix~\ref{ap}.
  The blue button at the bottom is a drop-down menu where the user picks the node-link
  layout algorithm.
  The green button is for rendering the network to the canvas and initializing further widgets.
  By hitting the green button, all the elements in this image is made disbled, except for the layout
  button.
  Further information is given in Section~\ref{use}.
  }\label{initial}
\end{figure}

There are multiple levels of representation entailed by the multilevel strategy.
This makes pertinent that one and only of the available levels is explicitly selected
so that the transformations are performed on that level.
Thus, there is always a selected level in which the tools perform actions, such
as link coloring or opening supernodes to expose their predecessor nodes.
As can be infered by the reader,
the number of nodes and links in each level which are visible are not fixed thoughout the
navigation.
The selection of the level to which the transformations should apply, and the number
of visible and total nodes and links in each level, are exposed in the table to the right
of the canvas, as in Figure~\ref{secondPhase}.
In the same table, the user may change the color and shape of the nodes in each layer of
a level, the color of the links of a level, and toggle show/hide of the links is a level,
by left/right clicks on the corresponding colored cell.

\begin{figure}[!h]\centering
 \includegraphics[width=\textwidth]{secondPhase}
  \caption{The elements in second step of the MlBiNetVis usage, dedicated to navigation on the network and transformation of the visualization.
  The user may change the layout used when supernodes are opened and the predecessors are then shown on the screen.
  Links may be shown or hided for all the levels by using the checkbutton.
  The text area assists in the output of information requested or useful for navigation.
  The canvas holds the node-link representations of the levels, which are transformed using the tools in the toolbar and the table in the right-side.
  Such table also makes evident key information about the status of the visualization.
  Further information is given in Section~\ref{use}.
  }\label{secondPhase}
\end{figure}

@@

The third group of buttons is more specific to the needs of the MlBiNetViz, and
are selected to make mouse operations entail special processes and transformations of the visualization.
The first button is selected for the interface to return core information from a clicked node:
id, level, number of predecessors of previous level, id of successor (if node is not on the last level),
number of neighbors (degree), sum of weight of links (strength).
Other information the analysts report necessity of are easily implemented, but currently only these informations
are returned to avoid unecessarily flooding the text area.
The second tool button is seleted to move nodes: specific nodes are individually moved if clicked and dragged;
a rectangular region is made available for moving all nodes therein if an empty region of the canvas
is clicked and the mouse cursor dragged before releasing the button.
The third tool button is selected for opening supernodes and exposing their predecessor nodes and links among them:
a region is defined by the user by clicking on an empty space on the canvas and gragging the mouse before releasing,
the supernodes therein are joined into a rectangular region that contains all of them,
and their predecessor nodes (which are on a finer level of detail) are drawn inside such rectangular region.
Open supernodes may be joined using the same tool, but it was found convenient to make available a
fourth button to assist in this task: if this tool is selected, one may join two open supernodes by
clicking on them in sequence.
The last tool button in this group is selected to perform three operations on open supernodes:
if an open supernode is clicked on and dragged, the rectangular shape is resized accordingly;
if there is no drag, i.e. the super node is clicked on and released, the links from it to the other
supernodes are attached to the next counterclockwise corner of its rectagular region.
When the open supernode is resized, the layout of its child nodes is optimized for the drawn region.
This is most useful when used in combination with the drag tool:
on the finer level of detail, the nodes are repositioned into a more compact disposition.
Then upon a (potentially minimal) supernode
risize, the child nodes are made to fit the whole rectangular area.
The last set of tools acts on the whole canvas:
changing the color of the background, zoom, left-right/up-down pan,
rotation of the entire network,
and toggling between current zoom and pan status and the initial setting.
The toolbar is contextualized in Figure~\ref{secondPhase}.

@@

This is a very brief account of the styles and software resources used to implement
MlBiNetVis. The interested reader is welcome to browse the software code and documentation
and contact the authors~\cite{mbnvcode}.

@@

\appendix

\section{Choice and fine-tuning of the coarsening algorithm}\label{ap}
The elements in Figure~\ref{initial}.
